Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-08-27 manuscript.qmd
put data-strat panels into one figure

These datasets allow us to investigate correlations between the involvement of individual LNLs, or between risk factors and involvement. In @fig-data-strat, we have plotted the prevalence of each contralateral LNL's involvement, stratified by T-category, number of ipsilaterally involved LNLs, and whether the tumor extended over the mid-sagittal line. A similar but more complete stratification is also tabulated in the appendix in @tbl-data-strat.

![Contralateral involvement stratified by T-category (left panel), the number of metastatic LNLs ipsilaterally (center panel), and whether the primary tumor extended over the mid-sagittal line or was clearly lateralized (right panel).](figures/fig_data_strat.svg){#fig-data-strat}

The left panel in @fig-data-strat indicates that T-category is correlated with contralateral involvement (as it is with overal involvement). This is simply because T-category may on average be considered a surrogate for the time between onset of disease and diagnosis. I.e., a patient with a T4 tumor was -- on average -- diagnosed later than a patient with a T1 tumor. Thus, the former did have more time to develop metastases.

@@

Lastly, the right panel in @fig-data-strat shows that patients with a tumor crossing the mid-sagittal line show contralateral involvement vastly more often compared to patients with clearly lateralized tumors. This makes intuitive sense, because the lymphatic system in the head and neck region is typically symmetric and thus no major vessels cross the midline. Therefore, interstitial fluids from the primary tumor -- which we assume to carry living malignant cells -- may only reach the blind-ended lymphatic vessels in the contralateral neck via short-ranged diffusion. Which in turn is only possible when the primary tumor is close enough to the mid-sagittal line or crosses it. <!-- citation(s) needed? or is this intuition good enough? -->

The three risk factors for contralateral involvement, midline extension, T-category, and ipsilateral involvement are correlated. For example, midline extension was present in 48% of advanced T-category tumors but only in 7% of early T-category tumors. The higher prevalence of contralateral metastases in advanced T-category tumors may thus be partially explained by a higher rate of midline extension in these tumors. However, @fig-data-strat-uncorr indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors. The figure considers only patients with lateralized tumors not extending over the midline. For early T-category and healthy ipsilateral levels I-V, only 1% (1 out of 84 patients) shows involvement of contralateral level II. This increases to 5% (9 out of 165 patients) if ipsilateral level II is involved, to 15% (9 out of 60 patients) if ipsilateral levels II and III are involved, and to 23% (6 out of 26 patients) for advanced T-category with ipsilateral levels II and III involved.<!-- TODO: make values here dynamic. -->

![Contralateral involvement prevalence by LNL for lateralized tumors. Shown are four different scenarios: 1) early T-category with an ipsilateral N0 neck (green), 2) early T-category with ipsilaterally LNL II involved (blue), 3) late T-category with ipsilateral LNL II metastatic (orange), and 4) early T-category with ipsilateral both levels II and III involved (red). This illustrates that even when controlling for the tumor's lateralization, T-category and ipsilaeral involvement are predictive of contralateral metastases.](figures/fig_data_strat_uncorr.svg){#fig-data-strat-uncorr}

>>>>

These datasets allow us to investigate correlations between the involvement of individual LNLs, or between risk factors and involvement. In @fig-data-strat, we have plotted the prevalence of each contralateral LNL's involvement, stratified by risk factors such as T-category, number of ipsilaterally involved LNLs, and whether the tumor extended over the mid-sagittal line. We also show in the fourth subfigre that T-category and ipsilateral involvement are not just rpedictive of contralateral involvement because they correlate with midline extension. A similar but more complete stratification is also tabulated in the appendix in @tbl-data-strat.

![Contralateral involvement stratified by T-category (top left panel), the number of metastatic LNLs ipsilaterally (top right panel), and whether the primary tumor extended over the mid-sagittal line or was clearly lateralized (bottom left panel). Additionally, in the bottom right, we show that even for lateralized tumors, T-category and ipsilateral involvement are correlated with contralateral metastasis by comparing the contralateral involvement prevalence for selected scenarios that vary in their T-category and ipsilateral involvement extent.](figures/fig_data_strat.svg){#fig-data-strat}

The top left panel in @fig-data-strat indicates that T-category is correlated with contralateral involvement (as it is with overal involvement). This is simply because T-category may on average be considered a surrogate for the time between onset of disease and diagnosis. I.e., a patient with a T4 tumor was -- on average -- diagnosed later than a patient with a T1 tumor. Thus, the former did have more time to develop metastases.

@@

The bottom left panel in @fig-data-strat shows that patients with a tumor crossing the mid-sagittal line show contralateral involvement vastly more often compared to patients with clearly lateralized tumors. This makes intuitive sense, because the lymphatic system in the head and neck region is typically symmetric and thus no major vessels cross the midline. Therefore, interstitial fluids from the primary tumor -- which we assume to carry living malignant cells -- may only reach the blind-ended lymphatic vessels in the contralateral neck via short-ranged diffusion. Which in turn is only possible when the primary tumor is close enough to the mid-sagittal line or crosses it. <!-- citation(s) needed? or is this intuition good enough? -->

The three risk factors for contralateral involvement, midline extension, T-category, and ipsilateral involvement are correlated. For example, midline extension was present in 48% of advanced T-category tumors but only in 7% of early T-category tumors. The higher prevalence of contralateral metastases in advanced T-category tumors may thus be partially explained by a higher rate of midline extension in these tumors. However, the bottom right panel of @fig-data-strat indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors. The figure considers only patients with lateralized tumors not extending over the midline. For early T-category and healthy ipsilateral levels I-V, only 1% (1 out of 84 patients) shows involvement of contralateral level II. This increases to 5% (9 out of 165 patients) if ipsilateral level II is involved, to 15% (9 out of 60 patients) if ipsilateral levels II and III are involved, and to 23% (6 out of 26 patients) for advanced T-category with ipsilateral levels II and III involved.<!-- TODO: make values here dynamic. -->
